---
title: "Student_dropout"
author: "Naro_KUOCH"
date: "2026-09-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(TimeDepFrail)
data(data_dropout)

head(data_dropout)
str(data_dropout)
```

CFUP: Standardized numerical covariate indicating the number of credits passed by the students by the end of the first semester.

time_to_event: Time (in semesters) at which a student decides to leave the university. A value greater than 6.0 indicates the student did not drop out during the follow-up (e.g. 6.1 semesters)

group: Categorical variable indicating the student's course of study, with 16 different levels from CosA, CosB, ... , CosP.

```{r}
unique(data_dropout$group)
```
```{r}
# check the number of missing value in each variable 
colSums(is.na(data_dropout))
```
```{r}
glimpse(data_dropout)
```
```{r}
data_dropout$time_to_event = as.numeric(data_dropout$time_to_event)
data_dropout$Gender = factor(data_dropout$Gender)
data_dropout$group  = factor(data_dropout$group)
tab=tbl_summary(data_dropout)
tab
```


# Event Indicator
if the student dropout the time_to_event <=6, otherwise the event din't drop in 6 semester
```{r}
#if dropped event = 1, event = 0 otherwise
data_dropout$event = ifelse( data_dropout$time_to_event <= 6, 1, 0)

```
# Check data set distribtion 
```{r}
# number of dropped out and didn't drop out in 6 semesters 
table(data_dropout$event)
```
```{r}
# number of male and female 
table(data_dropout$Gender)
```
```{r}
# number of student for each course
barplot(table(data_dropout$group))
```
```{r}

boxplot(
  data_dropout$CFUP,
  main = "Boxplot of CFUP",
  ylab = "CFUP"
)
```

# Estimate survival function with Kaplan-Meier
```{r}
library(survival)
library(ggsurvfit)

fit.KM <- survfit(Surv(time_to_event, event) ~ 1, data = data_dropout)
autoplot(fit.KM)
fit.KM
```
Probility of not dropping out always higher than 80%
```{r}
plot(fit.KM, fun = "F",
     xlab = "Semester",
     ylab = "Cumulated Probability of drop out")
```

Probability of dropping out never reach 50%

# Comparison between male and female 
```{r}
fit_gender <- survfit(Surv(time_to_event, event) ~ Gender, data = data_dropout)
autoplot(fit_gender)
fit_gender
```

```{r}
ggsurvfit(fit_gender)
```

# The logrank test
```{r}
survdiff(Surv(time_to_event, event) ~ Gender, data = data_dropout)
```
Since the p_value is less than 0.05 we can conclude that there is statistical significant of survival curves between male and female. 
# Comparison between courses
```{r}
fit_course <- survfit(Surv(time_to_event, event) ~ group, data = data_dropout)
fit_course
```
```{r}
ggsurvfit(fit_course)
```

# The logrank test
```{r}
survdiff(Surv(time_to_event, event) ~ group, data = data_dropout)
```
There is not statistical significant different between courses 

# Include courses in the comparison between gender
```{r}
survdiff(Surv(time_to_event, event) ~ Gender + strata(group), data = data_dropout)
```
```{r}
fitGC = survfit(
  Surv(time_to_event, event) ~ Gender + strata(group),
  data = data_dropout
)
ggsurvfit(fitGC)
```

Event we include the coures as confounding factor, the survival curves of males and females are stil statistically significant different 
# Include gender in the comparison between courses 
```{r}
survdiff(Surv(time_to_event, event) ~ group + strata(Gender), data = data_dropout)
```
Even if we add gender as cofounding factor, the difference of survival curves between coures are still not statistically significant 
# Effect of Academic performance on time to dropout with Cox model
```{r}
fit = coxph(Surv(time_to_event, event) ~ CFUP, data = data_dropout)
summary(fit)
```

The hazard ratio is statistically significant, and it is less than 1. Meaning that the increase of academic performance, decreases the risk of dropout.  

# Effect of Gender on time to dropout with Cox model
```{r}
summary(coxph(Surv(time_to_event, event) ~ Gender, data = data_dropout))
```

Being a male has higher risk to dropout than female. 
# Effect of course on time to dropout with Cox model
```{r}
summary(coxph(Surv(time_to_event, event) ~ group, data = data_dropout))
```
The harzard ratio is not statistically significant. 
cosB, cosC, cosE, cosF, cosO, and cosP have hazard ratio less than 1. Compare to cosA the students in those courses has less risk than CosA. While the other courses have hazard ratio more than 1 so they have more risk to dropout than the students in CosA.
# Effect of multiple covariates 
```{r}
fit_multi = coxph(Surv(time_to_event, event) ~ Gender + CFUP + group, data = data_dropout)
summary(fit_multi)
```
- GenderMale  has HR =  1.23756 meaning being in Male group has more risk to dropout than female group. 
- CFUP has HR = 0.27981 meaning that having passed high number of credits, has low risk to dropout. 
- groupCosB and cosE have HR < 1 meaning that students taking those courses have lower risk than students in groupCosA. 
- Other couses has HR >1 meaning that students in those courses has more risk to dropout than in course A
# Model Comparison
which model explain better Hazard ratio ? 
```{r}
step(fit_multi)
```
The model with all the combination of covariant has the lowest AIC.
# Residual Martigal
```{r}
data_dropout$residual <- residuals(fit_multi, type = "martingale")

par(mfrow = c(1, 3), mar = c(4.2, 2, 2, 2))
with(data_dropout, {

  plot(CFUP, residual)
  lines(lowess(CFUP, residual), lwd = 2)

  plot(residual ~ Gender)

  plot(residual ~ group)

})
```
The martingale residauls associated with variables Gendar and Goup have expected value around 0 but it is not the case for CFUP. Notice that there are outliers in those residuals 


# Validate proportional harzard with Schenfield Residaul 

```{r}
r = cox.zph(fit_multi)
r
```
Globally those variables are fine because the p value are large except the CFUP. 
```{r}
plot(r)
```
We saw that all the p_value are large except the one associated with CFUP. 
P_value <0.05 meaning the variable is not constant overtime which violated the assumption of cox model 
# Dealing with the time dependant covariate

we allow the CFUP to be time dependant
```{r}
cox.nonph = coxph(Surv(time_to_event, event) ~ Gender + CFUP + tt(CFUP)+ group, data = data_dropout, tt=function(x,t,...) x*t)
summary(cox.nonph)
```


